---
title: "dplyr and SQL"
author: "Randall Pruim"
date: "Big Data Ignite 2016"
output:
  beamer_presentation:
    colortheme: seagull
    fig_height: 3
    fig_width: 8
    fig_caption: false
    fonttheme: professionalfonts
    includes:
      in_header: ../header.tex
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, tidy=FALSE, size="small",
  cache = TRUE, cache.path = "cache/dplyr-sql-")
require(mosaic)
```


## dplyr and SQL integration

`dplyr` provides SQL integration

  * backends for several flavors of SQL exist 
    * Postgres, MySQL, MariaDB, BigQuery
  
  * others coming 
    * MS SQL server
  
## airlines data via MySQL (Smith College)

A data base that includes information on US flights since 1987 is available from
Smith College

```{r, message = FALSE}
require(dplyr)
airlinesdb <-
  src_mysql(
    "airlines", host = "scidb.smith.edu",
    user = "mth292", password = "RememberPi")
airlinesdb
```

## Connecting to the tables

```{r, warning = FALSE}
# supressing a couple warnings about type conversion
Airports <- tbl(airlinesdb, "airports")
Carriers <- tbl(airlinesdb, "carriers")
Flights <- tbl(airlinesdb, "flights")
Planes <- tbl(airlinesdb, "planes")
Weather <- tbl(airlinesdb, "weather")
```

## Taking a glimpse

```{r}
Weather %>% glimpse()
```

## NY Weather only

Weather data is only for NY airports in 2013.  Complete
24/7 data would include `r 24 * 365` records.

```{r}
Weather %>%
  group_by(origin, year) %>%
  summarise(num_records = n())
```

## Airports
```{r}
glimpse(Airports)
```

## GRR Airport

```{r, warning = FALSE}
Airports %>%
  filter(faa == "GRR")
```

## GRR Flights

```{r, eval = FALSE}
GRR2015 <-
  Flights %>%
  filter(year == 2015) %>% 
  filter(origin == "GRR" | dest=="GRR") %>%
  collect()
```

```{r, include = FALSE, eval = FALSE}
saveRDS(GRR2015, file = "GRR2015.rds")
```

```{r, include = FALSE}
GRR2015 <- readRDS("GRR2015.rds")
```

```{r}
nrow(GRR2015)
```

## Where do we go from here?

```{r}
GRR2015 <-
  GRR2015 %>%
  mutate(date = lubridate::mdy(paste(month, day, year))) 
GRR2015 %>%
  filter(origin == "GRR") %>%
  group_by(dest) %>%
  summarise(n = n()) %>%
  arrange(-n)
```
## Will my flight be on time?

```{r, warning = FALSE}
GRR2015 %>% 
  filter((day > 20 & month == 9) | 
         (day < 10 & month == 10)) %>%
  ggplot(aes(x = sched_dep_time, y = dep_delay)) +
  geom_point(alpha = 0.2) +
  geom_smooth()
```

## Will my flight be on time?

```{r, warning = FALSE, message = FALSE}
GRR2015 %>%
  filter(origin == "GRR") %>%
  ggplot(aes(x = date, y = dep_delay)) +
  geom_point(alpha = 0.2) + 
  geom_smooth() +
  ylim(0, 180) 
```
