---
title: "Introduction to dplyr"
author: "Randall Pruim"
date: "Big Data Ignite 2016"
output:
  beamer_presentation:
    colortheme: seagull
    fig_height: 3
    fig_width: 8
    fonttheme: professionalfonts
    includes:
      in_header: header.tex
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=FALSE, size="small",
                      fig.width=8, fig.height=6)
```

## Tidy (Rectangular) Data 

 * rows = cases/observational units
 * columns = variables  
 * no exceptions
 
### For plotting

 * row for each mark on the plot
 * columns contain information needed for position, color, size, shape, etc.


## Chaining 

The pipe syntax (`%>%`) takes a data frame (or data table) and sends it 
to the argument of a function.


- `x %>% f(y)` is the same as `f(x, y)`

- `y %>% f(x, ., z)` is the same as `f(x,y,z)`


## Building Tidy Data 

- object_name <- function_name(arguments) 
- object_name <- data_table %>% function_name(arguments) 
- object_name <-   
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;data_table %>%  
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; function_name(arguments) %>%  
  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; function_name(arguments)

- in chaining, the value (on left) %>% is **first argument** to the function (on right)


## 5 Main Data Verbs

Data verbs take data tables as input and give data tables as output

1. select(): subsets *variables* (and rename() )
2. filter():  subsets unwanted *cases*
3. mutate(): transforms the variable (and transmute() like mutate, returns only new variables)
4. arrange(): reorders the cases
5. summarize(): computes summary statistics

## Other Data Verbs
- distinct(): returns the unique values in a table
- sample_n(): take a random row(s)
- head():  grab the first few rows
- tail(): grab the last few rows
- group_by(): SUCCESSIVE functions are applied to groups
- ungroup():  reverse the grouping action
- summarise():  
    + min(), max(), mean(), sum(), sd(), median(), and IQR()
    + n(): number of observations in the current group
    + n_distinct(): number of unique values
    + first_value(), last_value(), and nth_value(x, n): (like x[1], x[length(x)], and x[n] )

## You should know that I did this:

```{r message=FALSE, warning=FALSE, echo=TRUE}
require(mosaic)
require(babynames)
require(NHANES)
Babynames <- babynames
names(NHANES) <- tolower(names(NHANES))
```

## Finally, some Examples! 

```{r}
Babynames %>% nrow()
Babynames %>% names()
```

## Finally, some Examples! 

```{r}
Babynames %>% glimpse()
```

## Finally, some Examples!

```{r}
Babynames %>% head()
```

## Finally, some Examples! 

```{r}
Babynames %>% sample_n(size=5)
```


## NHANES Data

```{r}
names(NHANES)
```

## 1. select(): subsets *variables*


find the sleep variables

```{r}
NHANESsleep <- 
  NHANES %>% 
  select(gender, age, weight, race1, race3, education, sleeptrouble,
         sleephrsnight, tvhrsday, tvhrsdaychild, physactive)

names(NHANESsleep)
dim(NHANESsleep)
```

## 2. filter():  subsets *cases* 

subset for college students

```{r fig.width=5, fig.height=3, fig.center=TRUE}
NHANESsleep <- NHANESsleep %>% filter(age %in% c(18:22))
histogram(~age, data=NHANESsleep)
```

## 3. mutate(): transforms the variable 

mutate or transmute to create a new variable?

```{r }
NHANESsleep %>% mutate(weightlb = weight*2.2) %>% head(3)
```

## 3. mutate(): transforms the variable 

mutate or transmute to create a new variable?

```{r }
NHANESsleep %>% transmute(weightlb = weight*2.2) %>% head(3)
```

## 5. summarize(): computes summary statistics

```{r}
# number of people (cases) in NHANES
NHANES %>% summarise(n())
```

## 5. summarize(): computes summary statistics

```{r}
# total weight of all the people in NHANES (silly)
NHANES %>% 
  mutate(weightlb = weight*2.2) %>% 
  summarise(sum(weightlb, na.rm=TRUE))
```

## 5. summarize(): computes summary statistics

```{r}
# mean weight of all the people in NHANES
NHANES %>% 
  mutate(weightlb = weight*2.2) %>% 
  summarise(mean(weightlb, na.rm=TRUE))
```

## 5. summarize() with group_by()

```{r}
# mean weight of all the people in NHANES, broken down by education
NHANES %>% 
  mutate(weightlb = weight*2.2) %>% 
  group_by(education) %>% 
  summarise(mean(weightlb, na.rm=TRUE))
```

## 4. arrange(): reorders the cases

```{r}
# mean weight, by education, sorted
NHANES %>% 
  mutate(weightlb = weight*2.2) %>% 
  group_by(education) %>% 
  summarise(avewt = mean(weightlb, na.rm=TRUE)) %>% 
  arrange(avewt)
```

## Your Turn 

When starting, it can be helpful to work with a small subset of the data.  When you have your data wrangling statements in working order, shift to the entire data table.

```{r}
SmallSubset <-
  Babynames %>%
  filter(year > 2000) %>%
  sample_n(size = 200)

names(SmallSubset)
```


## How many babies are represented?

```{r eval=FALSE}
SmallSubset %>%
  summarise(total = ????(n)) # a reduction verb
```

## How many babies are represented?

```{r}
SmallSubset %>%
  summarise(total = sum(n))
```


## How many babies are there in each year?

```{r eval=FALSE}
SmallSubset %>% 
  group_by(????) %>% 
  summarise(total = ????(n))
```

## How many babies are there in each year?

```{r}
SmallSubset %>% 
  group_by(year) %>% 
  summarise(total = sum(n))
```

## How many distinct names in each year?

```{r eval=FALSE}
SmallSubset %>%
  group_by(????) %>%
  summarise(name_count = n_distinct(????))
```

## How many distinct names in each year?

```{r}
SmallSubset %>%
  group_by(year) %>%
  summarise(name_count = n_distinct(name))
```

## How many distinct names of each sex in each year?

```{r eval=FALSE}
SmallSubset %>%
  group_by(????, ????) %>%
  summarise(????)
```

## How many distinct names of each sex in each year?

```{r}
temp <- SmallSubset %>%
  group_by(year, sex) %>%
  summarise(name_count=n_distinct(name))
data.frame(temp)
```


## Track the yearly number of Janes and Marys over the years.

```{r eval=FALSE}
Result <-
  Babynames %>%
  ????(name %in% c("Jane", "Mary")) %>% # just the Janes and Marys
  group_by(????, ????) %>% # for each year for each name
  summarise(count = ????)
```

## Track the yearly number of Janes and Marys over the years.

```{r}
Result <-
  Babynames %>%
  filter(name %in% c("Jane", "Mary")) %>% # just the Janes and Marys
  group_by(name, year) %>% # for each year for each name
  summarise(count = sum(n))
```

## Plot out the result

Put `year` on the x-axis and the count of each name on the y-axis.  Note that `ggplot()` commands use `+` rather than `%>%`.

```{r eval=FALSE}
ggplot(data=Result, aes(x = year, y = count)) +
  geom_point()
```

* *Map* the name (Mary or Jane) to the aesthetic of color.  Remember that mapping to aesthetics is always done inside the `aes()` function.
* Instead of using dots as the glyph, use a line that connects consecutive values: `geom_line()`.
* Change the y-axis label to "Yearly Births": `+ ylab("Yearly Births")`
* *Set* the line thickness to `size=2`. Remember that "setting" refers to adjusting the value of an aesthetic to a constant.  Thus, it's *outside* the `aes()` function.

## Plot out the result

```{r fig.width=5, fig.height=3, fig.center=TRUE}
ggplot(data=Result, aes(x = year, y = count)) +
  geom_line(aes(color=name), size=2) + 
  ylab("Yearly Births")
```


## Look at the *proportion* of births rather than the count

Filter first on female, then on Mary and Jane.  Meanwhile, calculate the proportion of names in the dataset which are Mary and Jane (as opposed to the proportion of all babies from the given year which is the *prop* variable).

```{r eval=FALSE}
Result2 <-
  Babynames %>%
  group_by(year) %>%
  mutate(total = ????(n)) %>%
  filter(????) %>%
  mutate(proportion = n / total)
```


## Look at the *proportion* of births rather than the count

Just as you did with count vs year, graph proportion vs year.
```{r eval=FALSE}
Result2 %>%
  Your ggplot statements go here!
```
* Add a vertical line to mark a year in which something happened that might relate to the increase or decrease the popularity of the name.  Example: The movie [*Whatever Happened to Baby Jane*](http://en.wikipedia.org/wiki/What_Ever_Happened_to_Baby_Jane%3F_%281962_film%29) came out in 1962.  The glyph is a vertical line: `geom_vline()`.

## Look at the *proportion* of births rather than the count

```{r}
Result2 <-
  Babynames %>%
  filter(sex == "F") %>%
  filter(sex %in% c("F"))  %>%
  group_by(year) %>%
  mutate(total = sum(n)) %>%
  filter(name %in% c("Mary", "Jane")) %>%
  mutate(proportion = n / total)
```

## Look at the *proportion* of births rather than the count

```{r fig.width=5, fig.height=3, fig.center=TRUE}
ggplot(data=Result2, aes(x=year, y=proportion)) +
  geom_line(aes(color=name), size=2) +
  ylab("Yearly Births") +
  geom_vline(xintercept=1962)
```

 
  
##  Pick out name(s) of interest to you 

Plot out their popularity over time.

##  Pick out name(s) of interest to you 

```{r fig.width=5, fig.height=3, fig.center=TRUE}
Frances <- babynames %>%
  filter(name== "Frances") 
```

##  Pick out name(s) of interest to you 

```{r fig.width=5, fig.height=3, fig.center=TRUE}
Frances %>% ggplot(aes(x=year, y=n)) +
  geom_point(aes(color=sex), size=.5) + 
  geom_line(aes(color=sex)) + 
  geom_vline(xintercept=2006) + scale_y_log10() +
  ylab("Yearly total on log10 scale")
```

## gather / spread (new verbs)

http://garrettgman.github.io/tidying/


-new_sp_m014 - new_rel_f65:  Counts of new TB cases recorded by group.   

-code for method of diagnosis (rel = relapse, sn = negative pulmonary smear, sp = positive pulmonary smear, ep = extrapulmonary)   
-code for gender (f = female, m = male)   
-code for age group (014 = 0-14 yrs of age, 1524 = 15-24 years of age, 2534 = 25 to 34 years of age, 3544 = 35 to 44 years of age, 4554 = 45 to 54 years of age, 5564 = 55 to 64 years of age, 65 = 65 years of age or older).  

```{r warning=FALSE, message=FALSE}
#install.packages(c("tidyr", "devtools"))
#devtools::install_github("garrettgman/DSR")
require(tidyr)
require(DSR)
data(who)
head(who)
```

## gather

```{r}
whoTidy <- who %>% gather("code", "value", 5:60)
whoTidy %>% sample_n(10)
```

## separate
```{r}
whoTidy <- whoTidy %>% separate(code, c("new", "var", "sexage"))
whoTidy %>% sample_n(10)
```

## separate

```{r}
whoTidy <- whoTidy %>% separate(sexage, c("sex", "age"), sep = 1)
whoTidy %>% sample_n(10)
```

## spread
```{r}
whoTidy <- whoTidy %>% spread(var, value)
whoTidy %>% sample_n(10)
```


## join (flights)
```{r warning=FALSE, message=FALSE}
require(nycflights13)
names(flights)
Flights <- flights
Airports <- airports
Airlines <- airlines
Weather <- weather
Planes <- planes
```

## join (flights)

```{r}
names(Flights)
names(Airports)
names(Airlines)
names(Weather)
names(Planes)
```

## join (flights)

```{r}
Flights %>% select(carrier, flight, tailnum, origin, dest) %>% head()
Airports %>% select(faa, name, lat, lon) %>% head()

Flights %>% 
  inner_join(Airports, by=c("dest" = "faa")) %>%
  select(carrier, flight, tailnum, origin, dest, name, lat, lon) %>%
  head()
```

## join (flights)

```{r}
# inner_join: return all rows from x where there are matching values in y, and all columns from x and y. 
Flights %>% inner_join(Airports, by=c("dest" = "faa")) %>% dim()

# left_join: return all rows from x, and all columns from x and y.
Flights %>% left_join(Airports, by=c("dest" = "faa")) %>% dim()

# right_join: return all rows from y, and all columns from x and y. 
Flights %>% right_join(Airports, by=c("dest" = "faa")) %>% dim()

# anti_join: return all rows from x where there are not matching values in y, keeping just columns from x.
Flights %>% anti_join(Airports, by=c("dest" = "faa")) %>% dim()

# full_join: return all rows and all columns from both x and y.
Flights %>% full_join(Airports, by=c("dest" = "faa")) %>% dim()
```


## lubridate

```{r warning=FALSE, message=FALSE}
require(lubridate)

rightnow <- now()
rightnow

day(rightnow)
week(rightnow)
month(rightnow, label=FALSE)
month(rightnow, label=TRUE)
year(rightnow)

minute(rightnow)
hour(rightnow)

yday(rightnow)
mday(rightnow)
wday(rightnow, label=FALSE)
wday(rightnow, label=TRUE)
```

## lubridate

Arithmetic with lubridate

```{r}
jan31 <- ymd("2013-01-31")
jan31 + months(0:11)

floor_date(jan31, "month") # round down to the nearest month
floor_date(jan31, "month") + months(0:11) + days(31)
jan31 + months(0:11) + days(31)

#Add and subtract months to a date without exceeding the last day of the new month 
jan31 %m+% months(0:11)
```

## parse_numeric
```{r message=FALSE, warning=FALSE}
require(tidyr)
require(readr)
parse_numeric("$1,200.34")
parse_numeric("-2%")

# The heuristic is not perfect - it won't fail for things that
# clearly aren't numbers
parse_numeric("-2-2")
parse_numeric("12abc34")
```

## humanparser

```{r}
require(humanparser)

```

## swirl

For a list of swirl courses, see: https://github.com/swirldev/swirl_courses

```
install.packages(“swirl”)
require(swirl) 
install_from_swirl("Getting and Cleaning Data")
swirl()
```
